\LARGE{An Example Responsible Machine Learning Workflow}\\\large{With Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-01-29 article-information-2019.tex
simulated results to appendix, update intro

%=================================================================
% Full title of the paper (Capitalized)
\Title{Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using measures with long-standing legal precedents: adverse impact ratio (AIR), marginal effect (ME), standardized mean difference (SMD), and additional straightforward group fairness measures. By combining interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to present a template workflow for important ML applications that require high accuracy and interpretability and that mitigate risks of discrimination.}

@@

\usepackage{caption}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds, i.e., ethics, social sciences, and technology.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is crystal clear: ML models, a common application of AI, have problems that responsible practitioners should likely attempt to remediate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. While this manuscript makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been resolved (even as singular entities, much less as complex intersectional phenomena), Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, in the form of interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools, to address a subset of these vexing problems for high-stakes, human-centered, or regulated ML applications.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

Section \ref{sec:m_and_m} describes methods and materials, including simulated and collected training datasets, interpretable and constrained model architectures, post-hoc explanations used to create an appealable decision-making framework, tests for DI and other social discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models, and post-hoc explanation and discrimination testing results are also presented for interpretable models. Section \ref{sec:disc} then discusses some nuances of the outlined modeling, explanation, and discrimination testing methods and results. Section \ref{sec:con} closes this manuscript with a brief summary of the outlined methods, materials, results, and discussion.

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
	\item \textbf{Notation}: spaces, datasets, \& models -- \textsection\ref{ssec:not}
	\item \textbf{Training data}: simulated data \& collected mortgage data -- \textsection\ref{ssec:sim_data}~and~\textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: AIR, ME, SMD and confusion matrix metrics -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository associated with Sections \ref{sec:m_and_m} and \ref{sec:res} -- \textsection\ref{ssec:soft}
\end{itemize}

To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulated Data}\label{ssec:sim_data}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mortgage Data}\label{ssec:mort_data}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulated Data Results}

\subsubsection{Constrained vs. Unconstrained Model Fit Assessment}\label{ssec: c_v_uc_sim}

\subsubsection{Interpretability and Post-hoc Explanation Results}\label{ssec:int_ex_sim}

\subsubsection{Discrimination Testing Results}\label{ssec:dis_sim}

\begin{table}[H]
	\captionsetup{width=7cm}
	\hskip-20pt\parbox{.48\textwidth}{
	\caption{Group size, accuracy, and FNR for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.} 
	\centering
	\begin{tabular}{ccccc}
		\toprule
		\textbf{Class}	& $\mathbf{N}$	& \textbf{Model} & \textbf{Accuracy} & \textbf{FNR} \\
		\midrule
		Protected 1	& 3,057		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.770 \\ 0.771\end{tabular} 	&  \begin{tabular}{@{}c@{}} 0.401 \\ 0.357\end{tabular} \\	
		\midrule			 
		Control 1		& 16,943		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.739 \\ 0.756\end{tabular}	& \begin{tabular}{@{}c@{}} 0.378 \\ 0.314\end{tabular} \\
		\midrule
		Protected 2	& 9,916		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.758 \\ 0.762\end{tabular}	&  \begin{tabular}{@{}c@{}} 0.331 \\ 0.302\end{tabular} \\	
		\midrule
		Control 2		& 10,084		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.729 \\ 0.756\end{tabular}	& \begin{tabular}{@{}c@{}} 0.420 \\ 0.332\end{tabular} \\		
		\bottomrule
		\label{tab:sim_n_acc}
	\end{tabular}
	}\hspace{10pt}
	\parbox{.48\textwidth}{
	\vspace{-13pt}
	\caption{AIR, ME, SMD, and FNR ratio for $g^\text{MGBM}$ and $g^\text{XNN}$ on the simulated test data.}
	\centering
	\begin{tabular}{ccccccc}
		\toprule
		\textbf{Model}		&  \begin{tabular}[x]{@{}c@{}}\textbf{Protected}\\\textbf{Class}\end{tabular} & \begin{tabular}[x]{@{}c@{}}\textbf{Control}\\\textbf{Class}\end{tabular}					& \textbf{AIR}									& \textbf{ME}									& \textbf{SMD} &  \begin{tabular}[x]{@{}c@{}}\textbf{FNR}\\\textbf{Ratio}\end{tabular}\\
		\midrule
		$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} 1 \\  2\end{tabular}		& \begin{tabular}{@{}c@{}} 1 \\ 2\end{tabular}	& \begin{tabular}{@{}c@{}} 0.752 \\ 1.10\end{tabular}	& \begin{tabular}{@{}c@{}} 9.7\% \\ -3.6\%\end{tabular}	& \begin{tabular}{@{}c@{}} -0.206 \\ 0.106\end{tabular}	& \begin{tabular}{@{}c@{}} 1.06 \\ 0.788\end{tabular}\\
		\midrule 
		$g^\text{XNN}$		& \begin{tabular}{@{}c@{}} 1 \\ 2\end{tabular}		& \begin{tabular}{@{}c@{}} 1 \\  2\end{tabular}	& \begin{tabular}{@{}c@{}} 0.727 \\ 0.976\end{tabular}	& \begin{tabular}{@{}c@{}} 12.0\%\\ 1.0\%\end{tabular}	& \begin{tabular}{@{}c@{}} -0.274 \\ 0.001\end{tabular}	& \begin{tabular}{@{}c@{}} 1.13 \\ 0.907\end{tabular} \\										
		\bottomrule
		\label{tab:sim_disc}
	\end{tabular}
	}
\end{table}

@@

\begin{figure}[H]
\centering
\includegraphics[width=8cm]{img/mort_mgbm_glob.png}
\caption{Global mean absolute Tree SHAP feature importance for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_mgbm_glob}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{img/mort_mgbm_glob_pdp_ice.png}
\caption{PD, ICE for 10 individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$, and histograms for the three most important input features of $g^\text{MGBM}$ on the mortgage test data.}
\label{fig:mort_mgbm_glob_pdp_ice}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{img/mort_mgbm_loc.png}
\caption{Tree SHAP values for three individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for the mortgage test data.}
\label{fig:mort_mgbm_loc}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=8cm]{img/mort_xnn_glob.png}
\caption{Global mean absolute Deep SHAP feature importance for $g^\text{XNN}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_xnn_glob}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{img/mort_xnn_glob_ridge.png}
\caption{\textbf{A.} Output layer $\gamma_k$ weights, \textbf{B.} corresponding $n_k$ ridge functions, and \textbf{C.} associated projection layer $\beta_j$ weights for $g^\text{XNN}$ on the mortgage test data.}
\label{fig:mort_xnn_glob_ridge}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{img/mort_xnn_loc.png}
\caption{Deep SHAP values for three individuals across selected percentiles of $g^\text{XNN}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_xnn_loc}
\end{figure}

@@

\begin{table}[H]
	\captionsetup{width=7cm}
	\hskip-20pt\parbox{.48\textwidth}{
		\caption{Group size, accuracy, and FNR for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.} 
		\centering
		\begin{tabular}{ccccc}
\toprule
\textbf{Class}	& $\mathbf{N}$	& \textbf{Model} 												& \textbf{Accuracy} 																			& \textbf{FNR} \\
\midrule
Black 		& 2,608		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.654 \\ 0.702\end{tabular} 				& \begin{tabular}{@{}c@{}} 0.457 \\ 0.308\end{tabular} \\	
\midrule			 
White 		& 28,361		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.817 \\ 0.857\end{tabular}			& \begin{tabular}{@{}c@{}} 0.508 \\ 0.360\end{tabular}\\		
\midrule
Female 		& 8,301		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}}  0.768 \\ 0.822\end{tabular}		& \begin{tabular}{@{}c@{}} 0.402 \\ 0.322\end{tabular}\\
\midrule
Male 		& 13,166		& \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.785 \\	0.847\end{tabular}			& \begin{tabular}{@{}c@{}} 0.497 \\ 0.347\end{tabular}\\
\bottomrule
			\label{tab:mort_n_acc}
		\end{tabular}
	}\hspace{10pt}
	\parbox{.48\textwidth}{
		\vspace{-13pt}
		\caption{AIR, ME, SMD, and FNR ratio for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.}
		\centering
		\begin{tabular}{ccccccc}
\toprule
\textbf{Model}		&  \begin{tabular}[x]{@{}c@{}}\textbf{Protected}\\\textbf{Class}\end{tabular} & \begin{tabular}[x]{@{}c@{}}\textbf{Control}\\\textbf{Class}\end{tabular}					& \textbf{AIR}									& \textbf{ME}									& \textbf{SMD} &  \begin{tabular}[x]{@{}c@{}}\textbf{FNR}\\\textbf{Ratio}\end{tabular}\\
\midrule
$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular}		& \begin{tabular}{@{}c@{}} White \\ Male\end{tabular}	& \begin{tabular}{@{}c@{}} 0.776 \\ 0.948\end{tabular}	& \begin{tabular}{@{}c@{}} 18.3\% \\ 4.1\%\end{tabular}	& \begin{tabular}{@{}c@{}} 0.628 \\ 0.084\end{tabular}	& \begin{tabular}{@{}c@{}} 0.900 \\ 0.810\end{tabular} \\
\midrule 
$g^\text{XNN}$		& \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular}		& \begin{tabular}{@{}c@{}} White \\ Male\end{tabular}	& \begin{tabular}{@{}c@{}} 0.743 \\ 0.955\end{tabular}	& \begin{tabular}{@{}c@{}} 21.4\% \\ 3.6\%\end{tabular}	& \begin{tabular}{@{}c@{}} 0.621 \\ 0.105\end{tabular}	& \begin{tabular}{@{}c@{}} 0.855 \\ 0.927\end{tabular} \\							
\bottomrule
			\label{tab:mort_disc}
		\end{tabular}
	}
\end{table}

@@

MGBM and XNN interpretable model architectures were selected for this text because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated for them to evaluate the constrained versions of these models. The same can be said of the presented explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML work flows and pipelines. While these approaches are promising responses to the black-box and discrimination problems in ML, they are just a small part of a burgeoning ecosystem of research and Python tools for responsible ML. Figure \ref{fig:blueprint} is a work flow blueprint that illustrates some of the additional steps that may be required to build a fully understandable and trustworthy ML system.\footnote{See: \href{https://github.com/jphall663/hc_ml}{Toward Responsible Machine Learning} for details regarding Figure \ref{fig:blueprint}.} While all the methods mentioned in Figure \ref{fig:blueprint} play an important role in increasing human trust and understanding of ML, a few pertinent references and Python resources are highlighted below as further reading.

\begin{figure}[H]
\centering
\includegraphics[width=8cm]{img/blueprint.png}
\caption{A diagram of a proposed holistic ML work flow in which interpretable models, post-hoc explanations, discrimination testing and remediation techniques, and other review and appeal mechanisms can create an understandable and trustworthy ML system.}
\label{fig:blueprint}
\end{figure}

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{\LARGE{An Example Responsible Machine Learning Workflow}\\\large{With Interpretable Models, Post-hoc Explanation, and Discrimination Testing}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using measures with long-standing legal precedents: adverse impact ratio (AIR), marginal effect (ME), standardized mean difference (SMD), and additional straightforward group fairness measures. By combining interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to present a template workflow for important ML applications that require high accuracy and interpretability and that mitigate risks of discrimination.}

@@

% for subtables
\usepackage{caption}
\usepackage{graphicx}
\usepackage{subcaption}

@@

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, present risks that responsible practitioners should likely attempt to mitigate. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that the interdependent issues of opaqueness, discrimination, or security vulnerabilities in ML have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures to address a subset of these vexing problems: interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available, free, and open source Python tools.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

Section \ref{sec:m_and_m} describes methods and materials, including collected training datasets, interpretable and constrained models, post-hoc explanations, tests for DI and other social discrimination, and public and open source software resources. In Section \ref{sec:res}, interpretable and constrained modeling results are compared to less interpretable and unconstrained models, and post-hoc explanation and discrimination testing results are also presented for interpretable models. Of course, an even wider array of tools and techniques are likely necessary to fully minimize discrimination, inaccuracy, privacy, and security risks associated with ML models. Section \ref{sec:disc} addresses the burgeoning Python ecosystem for responsible AI, along with appeal and override of automated decisions, and discrimination testing and remediation in practice. Section \ref{sec:con} closes this manuscript with a brief summary of the outlined methods, materials, results, and discussion.

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
	\item \textbf{Training data}: collected mortgage data -- \textsection\ref{ssec:mort_data}
	\item \textbf{ML models}: constrained, interpretable MGBM \& XNN models  -- \textsection\ref{ssec:mgbm}~and~\textsection\ref{ssec:xnn}
	\item \textbf{Post-hoc explanation techniques}: PD, ICE, \& Shapley values -- \textsection\ref{ssec:pd_ice}~and~\textsection\ref{ssec:shap}
	\item \textbf{Discrimination testing methods}: AIR, ME, SMD and confusion matrix metrics -- \textsection\ref{ssec:di}
	\item \textbf{Software resources}: GitHub repository associated with Sections \ref{sec:m_and_m} and \ref{sec:res} -- \textsection\ref{ssec:soft}
\end{itemize}

To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. The simulated data, based on the well-known Friedman datasets and with known feature importance and discrimination characteristics, is used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data is sourced from the Home Mortgage Disclosure Act (HMDA) database.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} Because unconstrained ML models, like gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), can be difficult to understand, trust, and appeal, even after the application of post-hoc explanation techniques, explanation analysis and discrimination testing are applied only to the constrained interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} and XNNs (\cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analysis. MGBM and XNN interpretable model architectures were selected for this text because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated for them to evaluate the constrained versions of these models. The same can be said of the presented explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML work flows and pipelines. While these approaches are promising responses to the black-box and discrimination problems in ML, they are just a small part of a burgeoning ecosystem of research and Python tools for responsible ML. Figure \ref{fig:blueprint} is a work flow blueprint that illustrates some of the additional steps that may be required to build a fully understandable and trustworthy ML system.

@@

\subsection{Mortgage Data}\label{ssec:mort_data}

@@

@@

\begin{figure}[H]
\centering
\includegraphics[width=9cm]{img/mort_mgbm_glob.png}
\caption{Global mean absolute Tree SHAP feature importance for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_mgbm_glob}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=16.5cm]{img/mort_mgbm_glob_pdp_ice.png}
\caption{PD, ICE for 10 individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$, and histograms for the three most important input features of $g^\text{MGBM}$ on the mortgage test data.}
\label{fig:mort_mgbm_glob_pdp_ice}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{img/mort_mgbm_loc.png}
\caption{Tree SHAP values for three individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for the mortgage test data.}
\label{fig:mort_mgbm_loc}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=9cm]{img/mort_xnn_glob.png}
\caption{Global mean absolute Deep SHAP feature importance for $g^\text{XNN}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_xnn_glob}
\end{figure}

@@

\begin{figure}[H]
\hskip-20pt\includegraphics[width=18cm]{img/mort_xnn_glob_ridge.png}
\caption{\textbf{A.} Output layer $\gamma_k$ weights, \textbf{B.} corresponding $n_k$ ridge functions, and \textbf{C.} associated projection layer $\beta_j$ weights for $g^\text{XNN}$ on the mortgage test data.}
\label{fig:mort_xnn_glob_ridge}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{img/mort_xnn_loc.png}
\caption{Deep SHAP values for three individuals across selected percentiles of $g^\text{XNN}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_xnn_loc}
\end{figure}

@@

\begin{table}[H]
	\hskip-10pt\begin{subtable}{0.45\textwidth}
	\captionsetup{width=7cm}
	\caption{Group size, accuracy, and FNR for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.} 
	\centering
	\resizebox{\columnwidth}{!}{%
	\begin{tabular}{ccccc}
		\toprule
		\textbf{Class} & $\mathbf{N}$ & \textbf{Model} & \textbf{Accuracy} & \textbf{FNR} \\
		\midrule
		Black & 2,608 & \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.654 \\ 0.702\end{tabular} & \begin{tabular}{@{}c@{}} 0.457 \\ 0.308\end{tabular} \\	
		\midrule			 
		White & 28,361 & \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular} & \begin{tabular}{@{}c@{}} 0.817 \\ 0.857\end{tabular} & \begin{tabular}{@{}c@{}} 0.508 \\ 0.360\end{tabular}\\		
		\midrule
		Female & 8,301 & \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular} & \begin{tabular}{@{}c@{}}  0.768 \\ 0.822\end{tabular} & \begin{tabular}{@{}c@{}} 0.402 \\ 0.322\end{tabular}\\
		\midrule
		Male & 13,166 & \begin{tabular}{@{}c@{}} $g^\text{MGBM}$ \\ $g^\text{XNN}$\end{tabular}	& \begin{tabular}{@{}c@{}} 0.785 \\ 0.847\end{tabular} & \begin{tabular}{@{}c@{}} 0.497 \\ 0.347\end{tabular}\\
		\bottomrule
		\label{tab:mort_n_acc}
	\end{tabular}%
	}
	\end{subtable}
	\hspace{10pt}
	\begin{subtable}{0.45\textwidth}
	\vspace{-14pt}
	\captionsetup{width=10cm}
	\caption{%
		\leftskip=45pt
		AIR, ME, SMD, and FNR ratio for $g^\text{MGBM}$ and $g^\text{XNN}$ on the mortgage test data.}
	\centering
	\scalebox{0.95}{
	\begin{tabular}{ccccccc}
		\toprule
		\textbf{Model} &  \begin{tabular}[x]{@{}c@{}}\textbf{Protected}\\\textbf{Class}\end{tabular} & \begin{tabular}[x]{@{}c@{}}\textbf{Control}\\\textbf{Class}\end{tabular} & \textbf{AIR} & \textbf{ME} & \textbf{SMD} & \begin{tabular}[x]{@{}c@{}}\textbf{FNR}\\\textbf{Ratio}\end{tabular}\\
		\midrule
		$g^\text{MGBM}$	& \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular}	& \begin{tabular}{@{}c@{}} White \\ Male\end{tabular} & \begin{tabular}{@{}c@{}} 0.776 \\ 0.948\end{tabular} & \begin{tabular}{@{}c@{}} 18.3\% \\ 4.1\%\end{tabular}	& \begin{tabular}{@{}c@{}} 0.628 \\ 0.084\end{tabular} & \begin{tabular}{@{}c@{}} 0.900 \\ 0.810\end{tabular} \\
		\midrule 
		$g^\text{XNN}$ & \begin{tabular}{@{}c@{}} Black \\ Female\end{tabular} & \begin{tabular}{@{}c@{}} White \\ Male\end{tabular} & \begin{tabular}{@{}c@{}} 0.743 \\ 0.955\end{tabular} & \begin{tabular}{@{}c@{}} 21.4\% \\ 3.6\%\end{tabular} & \begin{tabular}{@{}c@{}} 0.621 \\ 0.105\end{tabular} & \begin{tabular}{@{}c@{}} 0.855 \\ 0.927\end{tabular} \\							
		\bottomrule
		\label{tab:mort_disc}
	\end{tabular}
	}
	\end{subtable}
\captionsetup{width=10cm}
\caption{Discrimination measures for the mortgage test data.}
\end{table}

@@

Figure \ref{fig:blueprint} displays a holistic approach to ML model training, assessment, deployment meant to decrease discrimination, inaccuracy, privacy, and security risks for high-stakes, human-centered, or regulated ML applications.\footnote{See: \href{https://github.com/jphall663/hc_ml}{Toward Responsible Machine Learning} for details regarding Figure \ref{fig:blueprint}.}

\begin{figure}[H]
	\centering
	\includegraphics[width=11cm]{img/blueprint.png}
	\caption{An example responsible ML work flow in which interpretable models, post-hoc explanations, discrimination testing and remediation techniques, among other review and appeal mechanisms, can create an understandable and trustworthy ML system.}
	\label{fig:blueprint}
\end{figure}

While all the methods mentioned in Figure \ref{fig:blueprint} play an important role in increasing human trust and understanding of ML, a few pertinent references and Python resources are highlighted below as further reading. Any discussion of interpretable ML models would be incomplete without references to the seminal work of the Rudin group at Duke University and EBM or GA\textsuperscript{2}M models, pioneered by researchers at Microsoft and Cornell. In keeping with a major theme of this manuscript, models from these leading researchers and several other kinds of interpretable ML models are now available as open source Python packages. Among several types of currently available interpretable models, practitioners can now use Python to evaluate EBM in the \href{https://github.com/interpretml/interpret}{\texttt{interpret}} package, optimal sparse decision trees, GAMs in the \href{https://github.com/dswah/pyGAM}{\texttt{pyGAM}} package, a variant of Friedman's RuleFit  in the \href{https://github.com/scikit-learn-contrib/skope-rules}{\texttt{skope-rules}} package, monotonic calibrated interpolated lookup tables in \href{https://github.com/tensorflow/lattice}{\texttt{tensorflow/lattice}}, and \textit{this looks like that} interpretable deep learning \cite{osdt}, \cite{rulefit}, \cite{lattice}, \cite{this_looks_like_that}.\footnote{See: \href{https://github.com/xiyanghu/OSDT}{Optimal sparse decision trees}.}\textsuperscript{,}\footnote{See: \href{https://github.com/cfchen-duke/ProtoPNet}{\textit{This looks like that} interpretable deep learning}.} Additional, relevant references and Python functionality include:
